The Opioid Crisis: State Regulations and Labor Market Outcomes \\ [2ex] \large CEMFI

### 2024-06-18 slides/paper.tex
readme

For the past two decades, the United States has been facing the worst public health crisis in its recent history.
The so-called opioid epidemic, characterized by a rising number of deaths due to opioid overdose, has caused more than 680,000 fatalities between 1999 and 2022 (\cite{CDCMCD9920}, \cite{CDCMCD1822}), mostly among non-college-educated Americans.
The sheer magnitude of this crisis has significantly impacted the labor market, prompting economists to study the relationship between opioid misuse and labor market outcomes.
However, little is known about how labor market regulations have shaped this relationship.
Minimum wage rates emerge as a natural candidate for studying this dynamic since they affect the same subpopulation of low-educated workers and have increased with varying intensity during the same period.

@@

This paper argues that minimum wage rates influence the relationship between opioid misuse and labor market outcomes.
I measure this relationship using the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states, which require authorized users to check a patient's prescription history before prescribing opioids. 
These programs force addicts to either quit opioids or switch to illegal alternatives, with the choice depending on available labor market opportunities.
Quitting opioids, though difficult, improves health and job prospects.
However, in areas with high (binding) minimum wages, reduced labor demand might lower the benefits of quitting.
My estimations show increases in labor force participation and unemployment rates only where the minimum wage was not binding, while a surge in heroin consumption occurred where it was.

%at the time the law is passed

% Condensate the whole background part into a single (two) paragraph maybe not mention lit yet

These findings contribute to the growing body of literature studying the economic effects of the opioid crisis.
Within this literature, the origins of the epidemic have been widely discussed, with a supply-side view of its origins now largely accepted over a demand-side explanation.
This perspective has led economists to focus on supply-side interventions to explore the causal relationship between opioid consumption and labor market outcomes.
While well-intentioned, many of these policies have had unintended consequences, such as the rise in heroin consumption following the reformulation of OxyContin.
However, Prescription Drug Monitoring Programs (PDMPs), have been less scrutinized.
Recent research indicates that, in their more stringent form, these programs can effectively reduce prescription opioid misuse.

\begin{comment}
    These findings contribute to the growing body of literature studying the economic effects of the opioid crisis and of the policies implemented to curb it.
    Within this literature, the origins of the epidemic have been widely discussed, with a supply-side view of its origins now largely accepted over a demand-side explanation.
    This perspective has led economists to focus on supply-side interventions to explore the causal relationship between opioid consumption and labor market outcomes.
    Prescription Drug Monitoring Programs (PDMPs) 
    Recent research indicates that, in their more stringent form, these programs can effectively reduce prescription opioid misuse.
\end{comment}

Yet, how heterogenous these effects are across labor markets remains an open question.
Different incentives can lead victims of opioid misuse to take different decisions regarding both their health and their economic prospects. 
Labor market regulations, by altering the labor outcomes of workers, influence this decision-making process by affecting the incentives to quit opioids -- a family of drugs that can severely impair the ability to work.
To understand how these regulations have shaped the ongoing crisis, I focus on minimum wage rates, which have increased in all states at different speeds during the same period as the opioid crisis, thereby affecting the labor market opportunities of low-educated workers.
Critical to our argument are not the rates themselves, but how binding they are across different labor markets: the more binding the minimum wage, the less incentives workers have to quit opioids, as they face a tougher job market search.

I use several data sources to obtain a distribution of coefficients of the impact of opioid misuse on labor market outcomes across different minimum wage bindingness levels.
The US Bureau of Labor Statistics (BLS) LAUS database provides county-level labor market statistics, while the US Census Bureau's Population Estimates Program (PEP) offers county-level population and demographic data.
From \textcite{Horwitz2021} I obtain data on Prescription Drug Monitoring Programs (PDMPs). 
The BLS's Occupational Employment and Wage Statistics (OEWS) database provides national industry-specific occupations and wages statistics, \textcite{Eckert2020} offers county-level industry-specific employment data, and the US Department of Labor's Changes in Basic Minimum Wages provides information on minimum wage rates; all of these data are relevant for constructing a measure of minimum wage bindingness.

The Centers for Disease Control (CDC) National Vital Statistics System offers data on total drug poisoning mortality by county, while the CDC's Multiple Cause of Death database provides overdose deaths at the state-year level for different drugs.
Finally, the Drug Enforcement Administration's Automation of Reports and Consolidated Orders System (ARCOS) offers opioid distribution data.
The joint panel covers the United states for the period 2003-2016.

The results are striking in several dimensions.
After constructing a measure of minimum wage bindingness known as the Kaitz-\(\rho\) index, I find that the impact of prescription policies on labor market outcomes is indeed heterogeneous across different levels of minimum wage bindingness.
First, I find that policies aimed at stopping opioid misuse increased participation rates only where the minimum wage was not binding.
Secondly, the increase was short-lived (less than two years), but the impact on unemployment rates was persistent.
Moreover, their impact on prescription rates seem to be the main, but not the only, driver of these results.
Finally, prescription policies had no labor market consequences where where the minimum wage was binding, but their impact on heroin deaths was increasing in minimum wage bindingness.

Therefore, this paper sheds light on how labor market regulations, and in particular minimum wage rates, give rise to heterogeneities in the effects of the opioid crisis on the labor market, contributing to a better understanding of the ongoing crisis and the policies implemented to address it.
More broadly, it contributes to the understanding of how labor market regulations interact with drug disorders.

The paper is structured as follows.
First, it provides a summary of the opioid crisis and the effects it has had on the labor market as uncovered by the literature.
Then, it discusses some of the policies implemented to curb the opioid misuse and its effectiveness.
Later, I describe how different minimum wage rates can explain some of these policy consequences and the heterogeneity of the labor market outcomes. 
A model is introduced to formalize the intuition behind this argument.
The employed data is discussed before presenting the empirical strategy and obtained results.

@@

The following model formalizes the intuition outlined above. 
An addicted worker \(i\) faces a health/performance cost of opioid consumption \(O_{it}\) at time \(t\).
This cost depends on whether she consumes prescription/legal opioids \(l_{it}\) or illegal opioids \(h_{it}\) (e.g., heroin).
The type of opioid consumed depends on the availability of legal opioids \(z_{it}\), which equals 1 if an effective prescription law is in place.
I assume that opioid misuse impedes the individual's ability to work.

@@

This section presents the main results of the paper. 
I start describing the construction of the Kaitz-\(\rho\) index, which is used to measure the bindingness of the minimum wage.
Then, I present the main especification of the paper, which is employed to estimate the effects of Prescription Drug Monitoring Programs on labor force participation and unemployment rates.
The coefficients of the regression are compared across the spatial distribution of minimum wage bindingness measures.
Finally, I repeat the same exercise using overdose deaths for different opioids as the dependent variable.

@@

The argument outlined above hinges on the assumption that the minimum wage can affect the labor market opportunities of low-educated workers.
In other words, at least in some areas it must be the case that the minimum wage is an effective, or binding, constraint on the labor market.
A common approach to measure this bindingness is through the Kaitz-\(\rho\) index, which can be defined as

@@

where \(w^{\text{\textit{min}}}_{s(c)t}\) is the minimum wage in the state where the county \(c\) is located at time \(t\), and \(w^{\rho}_{ct}\) is the \(\rho\) percentile of the wage distribution.
This measure generates an spatial distribution of the bindingness of the minimum wage, which can be used to compare the effects of the opioid crisis across different labor markets.
Figure 1 shows the distribution of the Kaitz-\(\rho\) index across counties in the United States for the year a Must Query PDMP is implemented in their states, for each of the different percentiles of the wage distribution included in the panel.
For the main results, I use the 10th percentile, but the findings are similar across the distribution -- the counties for which the minimum wage is more binding with respect to lower percentiles tend to be the same as those for which it is more binding with respect to higher percentiles.

The indices are obtained using the data on the wage distribution at the occupation-industry level, and the shares of employment of each industry in each county. 
In particular, let \(w^{\rho}_{oit}\) be the percentile \(\rho\) of the wages at the occupation \(o\) in the industry \(i\) at time \(t\). 
Then

@@

where \(e_{oi}\) is the number of workers in the occupation \(o\) and \(E_{it}\) is the total number of workers in the industry \(i\) at time \(t\).
That is, by weighting the wage percentiles by the share of employment of each occupation in the industry, I obtain the wage percentile for the industry.
After that, I compute

@@

where \(e_{ci}\) is the number of workers in the industry \(i\) in the county \(c\) and \(E_c\) is the total number of workers in the county.
\(w^{\rho}_{ct}\) is the wage percentile for the county.
Finally, the minimum wage rates at the state level are already included in the panel.

@@

The main specification computes individual-level effects following \textcite{Arkhangelsky2024} and has the form

@@

Here, \(y_{ct}\) is a labor market outcome, either the labor force participation rate or the unemployment rate, \(\alpha_c\) are county fixed effects, and \(\lambda_t\) are time fixed (year-month) effects.
Relative time is represented with \(k = t - z_c\), the time difference of the calendar time from the implementation of a PDMP in the state, represented by \(z_s\).
\(\tau_{ck}\) are the coefficients of interest and \(\nu_{ct}\) is an idiosyncratic shock.

The covariates \(X_{ct}\) include demographic shares (race, origin, sex and age), sector shares of employment (at the 2-digits NAICS codes), and total drug poisoning mortality, which might be a relevant variable in policy-making.
Crucial for identification, I assume that the state-level rollout of the PDMPs is exogenous at the county level after conditioning on the covariates.
Moreover, this strategy requires that there is no anticipation of the policy up to \(h\) periods. 
I assume no anticipation whatsoever, so that the coefficients are identified from the very start of the policy.
In this context, this means that no addicted individual decided to look for a job or switch to heroin because of the expected implementation of the PDMP.

After running (10), a distribution of individual (county-level) coefficients is obtained.
These coefficients are then compared across the spatial distribution of the Kaitz-\(\rho\) index at the moment of treatment.
The comparison is made by running a binscatter following \textcite{Cattaneo2024}, which allows for an optimal binning of the data.

@@

Using the labor force participation rate as the dependent variable, the results of the main specification are displayed in Figure 2.
The figure shows the distribution of the coefficients of the impact of the passing of the Must Query PDMPs on labor force participation rates across different levels of minimum wage bindingness, for different periods relative to treatment.
In the upper right panel we can see the immediate impact of the policy.
For those counties where the minimum wage is less binding, the effect is positive (1.37 p.p.) and significant (0.513, 2.22) at conventional levels.
The effects become smaller and more uncertain as the minimum wage becomes more binding.
For those counties in which the minimum wage has a higher impact on labor demand, the point estimate becomes negative (-0.693) but not significant (-2.54, 1.16).

The effect seems to be persistent in the short run (upper right and lower right panels), but fades away after two years (lower right)\footnote{The confidence interval of the most binnding counties is not displayed in this last panel for visualization purposes. The values are (-3.10, 1.90)}.
Figure 3 shows the time averages of the coefficients for two years after the law is passed.
The results remain positive and significant for those counties in which the minimum wage is less binding (1.06 p.p., (0.584, 1.55)), while they are not significant for those where the minimum wage was more binding.

The results seem to be driven mostly by the impact of the policy on prescription rates.
Although the overlap between the passing of the policies (Table 1) and the data for prescription rates (2006-2014) is scarce, we can still obtain som estimates on the impact of Must Query PDMPs on prescription rates per capita.
Figure 4 runs the model (10) with the prescription rates per capita as the dependent variable.
The two months after the treatment is passed are displayed.
We see that the policy is able to decrease the amount of prescriptions, with the biggest drops occuring where the minimum wage is the least and the most binding (the effects in these areas are not statistically different). 
Although the target of these policies where not to decrease prescriptions \textit{per se}, but only tackling misuse cases, the results suggest that the policies are effective in reducing the amount of opioids prescribed on the aggregate.

Moreover, this seems to be the main channel through which the policies affect labor market outcomes: Figure 8 in the appendix shows that the relation between labor market outcomes and minimum wage bindingness is not significant when controlling for prescription rates.
Figures 9 and 10 in the Appendix show the the point estimates of the binscatter for different Kaitz indices, for six months and two years after the policy is passed.
It is easy to see that the results do not change much across different measures, this being especially true the closer in time the treatment was implemented.

@@

The analysis is repeated using the unemployment rate as the dependent variable.
The results are displayed in the same fashion in Figure 5.
The distribution of the coefficients reveal that almost all the positive effect it was found in the labor force participation rates translated into an increase in the unemployment rate.
For those counties where the minimum wage is less binding, the effect is positive (1.02 p.p.) and significant (1.46, 0.575) at conventional levels. 
The point estimate is only 0.35 p.p. below the one it was found for the labor force participation rate.
Although a positive effect is also found for those counties where the minimum wage is more binding (0.640 p.p.), the effect is not significant (-0.0252, 1.31).

These results shouldn't be surprising, at least in the short run: if people with severe opioid addictions start looking for a job after the implementation of a PDMP, they might not be able to find one immediately even in those counties where the minimum wage is less binding.
Strikingly though, the impact on unemployment persists after two years.
The lower right panel shows that, even after the increase in labor force participation rates fades away, the increase in unemployment rates remains high (2.77 p.p.) and significant (1.87, 3.68) where the minimum wage is less binding.
Whether the causes of this persistence are related to opioid consumption (through, say, relapsing) or to characterisctics of the workers or the labor market (say, if these workers face more frictional unemployment) remains an open question.

@@

The time average of the coefficients, displayed in Figure 6, confirm our previous result.
The effect of the policy on unemployment rates is positive and significant for those counties where the minimum wage is less binding (1.25 p.p., (0.97, 1.52)), while it is not significant for those where the minimum wage is more binding (0.025 p.p., (-0.11, 0.16)).
Again, the results seem to be driven by the impact of the policy on prescription rates, especially in the medium run: Figure 11 in the Appendix shows that the impact stops being significantly different from zero after one year once we control for prescriptions per capita.
Figures 12 and 13 in the Appendix show the binscatter for different Kaitz indices, for six months and two years after the policy is passed.
The results do not change substantially across different measures.

@@

Finally, and in accordance with the model outlined above, I show some evidence that the rollout of Must Query PDMPs has had a positive impact on the consumption of illegal opioids in those counties where the minimum wage was more binding.
For this purpose, I run the same estimation strategy as before, but using the number of overdose deaths per capita for different opioids as the dependent variable. 
The data is at the state-year level, and the number of total poisoning deaths is not used as a control variable.
Figure 7 displays the scatter plot of the time average of the coefficients on the Kaitz-0.10 index of the states, for the two years after the policy is passed.
The opioids included are heroin, methadone, other opioids and other synthetic narcotics, following the ICD-10 classification.

The coefficients are regressed on the Kaitz index and a constant. 
Tables 2-5 show the values of the regression coefficients for each of the different opioids.
Although admittedly weaker than the evidence presented in previous sections due to data constaints, it is interesting that the only marginally significant relation is found for heroin.
This suggest that the positive relation between black market switching and low labor demand is present, but we cannot assert this relationship with the same degree of certainty.

It seems implausible that part of the effect is absorbed by the consumption of other opioids.
Methadone is often used as a treatment for opioid adiction and can be legally used under a doctor's supervision.
Other opioids include drugs such as morphine and codeine, but also semi-synthetic opioid analgesics (e.g. drugs such as oxycodone, hydrocodone, hydromorphone, and oxymorphone), some of them targeted by the PDMPs.
Finally, although the category of other synthetic narcotics includes fentanyl and its analogues, its use didn't used to be as widespread as it is today during the period considered.

@@

This paper contributes to the understanding of how labor market regulations, particularly minimum wage rates, shape the relationship between opioid misuse and labor market outcomes. By leveraging the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states, this study provides a nuanced view of how labor market opportunities influence the incentives for opioid users to quit or switch to illegal alternatives.

The findings reveal that labor force participation and unemployment rates increased as a result of prescription policies only in areas where the minimum wage was not binding. In contrast, a surge in heroin consumption occurred where the minimum wage was binding, highlighting the complex interplay between labor market conditions and opioid misuse. These results underscore the importance of considering labor market regulations when evaluating the economic impacts of the opioid crisis.

Furthermore, this study adds to the growing body of literature that examines the economic effects of the opioid epidemic. While previous research has primarily focused on the direct consequences of opioid misuse on labor market outcomes, this paper emphasizes the role of labor market regulations in mediating these effects. The heterogeneity in the impact of prescription policies across different levels of minimum wage bindingness suggests that policies aimed at curbing opioid misuse must account for the broader economic context to be effective.

In summary, the analysis demonstrates that higher minimum wages, by reducing labor demand for low-skilled workers, can diminish the incentives for opioid users to quit, thereby exacerbating the shift towards illegal alternatives like heroin. These insights highlight the need for a comprehensive approach to addressing the opioid crisis, one that integrates labor market policies with public health interventions to effectively mitigate the epidemic's adverse effects on both health and economic outcomes.

Future research should continue to explore the interplay between labor market regulations and health crises, considering the potential unintended consequences of well-intentioned policies. By doing so, policymakers can design more effective interventions that address the root causes of public health crises while also supporting economic stability and growth.

[MOVING FORWARD]

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_lab_for_rate.png}
    \caption{}
    \label{fig:lab_for_rate}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_lab_for_rate_ta.png}
    \caption{}
    \label{fig:lab_for_rate_ta}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_prescriptions.png}
    \caption{}
    \label{fig:prescriptions}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_unemp_rate.png}
    \caption{}
    \label{fig:unemp_rate}
\end{figure}

>>>>

% CHECKED
For the past two decades, the United States has been facing the worst public health crisis in its recent history.
The so-called opioid epidemic, characterized by a rising number of deaths due to opioid overdose, has caused more than 680,000 fatalities between 1999 and 2022 (\cite{CDCMCD9920}, \cite{CDCMCD1822}), mostly among non-college-educated Americans.
The sheer magnitude of this crisis has significantly impacted the labor market, prompting economists to study the relationship between opioid misuse and labor market outcomes.
However, little is known about how labor market regulations have shaped this relationship.
Minimum wage rates emerge as a natural candidate for studying this dynamic since they affect the same subpopulation of low-educated workers and have increased with varying intensity during the same period.

@@

% CHECKED
This paper argues that minimum wage rates influence the relationship between opioid misuse and labor market outcomes.
I measure this relationship using the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states, which require authorized users to check a patient's prescription history before prescribing opioids. 
These programs force addicts to either quit opioids or switch to illegal alternatives, with the choice depending on available labor market opportunities.
Quitting opioids, though difficult, improves health and job prospects.
However, in areas with high (binding) minimum wages, reduced labor demand might lower the benefits of quitting.
My estimations show increases in labor force participation and unemployment rates only where the minimum wage was not binding, while a surge in heroin consumption occurred where it was.

% CHECKED
These findings contribute to the growing body of literature studying the economic effects of the opioid crisis and the policies implemented to curb it.
Within this literature, the origins of the epidemic have been widely discussed, with a supply-side view of its origins now largely accepted over a demand-side explanation (\cite{Ruhm2018}).
This perspective has led economists to focus on supply-side interventions to explore the causal relationship between opioid consumption and labor market outcomes.
While well-intentioned, many of these policies have had unintended consequences, such as the rise in heroin consumption following the reformulation of OxyContin (\cite{Alpert2018}).
However, Prescription Drug Monitoring Programs (PDMPs) have been less scrutinized.
Recent research indicates that, in their more stringent form, these programs can effectively reduce prescription opioid misuse (\cite{Buchmueller2018}).

% CHECKED
Yet, how heterogenous these effects are across labor markets remains an open question.
Different incentives can lead victims of opioid misuse to take different decisions regarding both their health and their economic prospects. 
Labor market regulations, by altering the labor outcomes of workers, influence this decision-making process by affecting the incentives to quit opioids -- a family of drugs that can severely impair the ability to work.
To understand how these regulations have shaped the ongoing crisis, I focus on minimum wage rates, which have increased in all states at different speeds during the same period as the opioid crisis, thereby affecting the labor market opportunities of low-educated workers.
Critical to our argument are not the rates themselves, but how binding they are across different labor markets: the more binding the minimum wage, the less incentives workers have to quit opioids, as they face a tougher job market search.

% CHECKED
I use several data sources to obtain a distribution of coefficients for the impact of opioid misuse on labor market outcomes across different levels of minimum wage bindingness.
The US Bureau of Labor Statistics (BLS) LAUS database provides county-level labor market statistics, while the US Census Bureau's Population Estimates Program (PEP) offers county-level population and demographic data.
Data on Prescription Drug Monitoring Programs (PDMPs) is obtained from \textcite{Horwitz2021}. 
The BLS's Occupational Employment and Wage Statistics (OEWS) database provides national industry-specific occupations and wage statistics. 
County-level industry-specific employment data comes from \textcite{Eckert2020}, and information on minimum wage rates is sourced from the US Department of Labor's Changes in Basic Minimum Wages.
These last three datasets are crucial for constructing a measure of minimum wage bindingness.

% CHECKED
The Centers for Disease Control (CDC) National Vital Statistics System offers data on total drug poisoning mortality by county, while the CDC's Multiple Cause of Death database provides overdose deaths at the state-year level for different drugs.
Finally, the Drug Enforcement Administration's Automation of Reports and Consolidated Orders System (ARCOS) offers opioid distribution data.
The joint panel covers the United states for the period 2003-2016.

% CHECKED
The results are striking in several dimensions.
After constructing a measure of minimum wage bindingness known as the Kaitz-\(\rho\) index, I find that the impact of prescription policies on labor market outcomes is indeed heterogeneous across different levels of minimum wage bindingness.
First, policies aimed at stopping opioid misuse increased labor force participation rates only where the minimum wage was not binding. 
Secondly, the increase was short-lived (less than two years), but the impact on unemployment rates was persistent.
Moreover, the impact on prescription rates appears to be the main, but not the only, driver of these results.
Finally, prescription policies had no labor market consequences where the minimum wage was binding, but their impact on heroin deaths increased with minimum wage bindingness.

% CHECKED
Therefore, this paper sheds light on how labor market regulations, particularly minimum wage rates, create heterogeneities in the effects of the opioid crisis on the labor market, contributing to a better understanding of the ongoing crisis and the policies implemented to address it.
More broadly, it contributes to understanding how labor market regulations interact with drug disorders.

% CHECKED
The paper is structured as follows:
First, it provides a summary of the opioid crisis and its effects on the labor market as uncovered by the literature.
Then, it discusses some of the policies implemented to curb opioid misuse and their effectiveness.
Later, it describes how different minimum wage rates can explain some of these policy consequences and the heterogeneity of labor market outcomes. 
A model is introduced to formalize the intuition behind this argument.
The employed data is then discussed, followed by the empirical strategy and obtained results.

@@

% CHECKED (the whole section)
The following model formalizes the intuition outlined above. 
An addicted worker \(i\) faces a health/performance cost of opioid consumption \(O_{it}\) at time \(t\).
This cost depends on whether she consumes prescription/legal opioids \(l_{it}\) or illegal opioids \(h_{it}\) (e.g., heroin).
The type of opioid consumed depends on the availability of legal opioids \(z_{it}\), which equals 1 if an effective prescription law is in place.
I assume that opioid misuse impedes the individual's ability to work.

@@

% CHECKED
This section presents the main results of the paper. 
First, I describe the construction of the Kaitz-\(\rho\) index, which measures the bindingness of the minimum wage.
Next, I introduce the primary specification used to estimate the effects of Prescription Drug Monitoring Programs (PDMPs) on labor force participation and unemployment rates. 
The regression coefficients are then compared across the spatial distribution of minimum wage bindingness measures.
Finally, I repeat the analysis using overdose deaths for different opioids as the dependent variable.

@@

% CHECKED
The argument outlined above hinges on the assumption that the minimum wage can affect the labor market opportunities of low-educated workers.
In other words, at least in some areas, the minimum wage must act as an effective, or binding, constraint on the labor market.
A common approach to measure this bindingness is through the Kaitz-\(\rho\) index, which can be defined as:

@@

% CHECKED
where \(w^{\text{\textit{min}}}_{s(c)t}\) is the minimum wage in the state where the county \(c\) is located at time \(t\), and \(w^{\rho}_{ct}\) is the \(\rho\) percentile of the wage distribution.
This measure generates an spatial distribution of the bindingness of the minimum wage, which can be used to compare the effects of the opioid crisis across different labor markets.
Figure 1 shows the distribution of the Kaitz-\(\rho\) index across counties in the United States for the year a Must Query PDMP is implemented in their states, for each of the different percentiles of the wage distribution included in the panel.
For the main results, I use the 10th percentile, but the findings are similar across the distribution -- the counties where the minimum wage is more binding at lower percentiles tend to be the same as those where it is more binding at higher percentiles.

% CHECKED
The indices are obtained using data on the wage distribution at the occupation-industry level and the employment shares of each industry in each county. 
Specifically, let \(w^{\rho}_{oit}\) be the \(\rho\) percentile of the wages in occupation \(o\) within industry \(i\) at time \(t\). 
Then,

@@

% CHECKED
where \(e_{oi}\) is the number of workers in occupation \(o\) and \(E_{it}\) is the total number of workers in industry \(i\) at time \(t\).
That is, by weighting the wage percentiles by the share of employment in each occupation within the industry, I obtain the wage percentile for the industry.
Next, I compute

@@

% CHECKED
where \(e_{ci}\) is the number of workers in industry \(i\) in county \(c\) and \(E_c\) is the total number of workers in the county.
\(w^{\rho}_{ct}\) represents the wage percentile for the county.
The state-level minimum wage rates are already included in the panel.

@@

% CHECKED
The main specification computes individual-level effects following \textcite{Arkhangelsky2024} and is given by

@@

% CHECKED
where \(y_{ct}\) is a labor market outcome, either the labor force participation rate or the unemployment rate, \(\alpha_c\) are county fixed effects, and \(\lambda_t\) are time fixed effects (year-month).
Relative time is represented by \(k = t - z_c\), the time difference from the implementation of a PDMP in the state (\(z_s\)).
\(\tau_{ck}\) are the coefficients of interest and \(\nu_{ct}\) is an idiosyncratic shock.

% CHECKED
The covariates \(X_{ct}\) include demographic shares (race, origin, sex and age), sector shares of employment (at the 2-digits NAICS code level), and total drug poisoning mortality, which might be a relevant variable in policy-making.
Crucial for identification, I assume that the state-level rollout of the PDMPs is exogenous at the county level after conditioning on the covariates.
Additionally, this strategy requires that there is no anticipation of the policy up to \(h\) periods. 
I assume no anticipation whatsoever, so that the coefficients are identified from the very start of the policy.
In other words, no addicted individual decided to look for a job or switch to heroin due to the expected implementation of the PDMP.

% CHECKED
After running (10), a distribution of individual (county-level) coefficients is obtained.
These coefficients are then compared across the spatial distribution of the Kaitz-\(\rho\) index at the moment of treatment.
The comparison is made by running a binscatter following \textcite{Cattaneo2024}, which allows for an optimal binning of the data.

@@

% CHECKED
Using the labor force participation rate as the dependent variable, the results of the main specification are displayed in Figure 2.
The figure shows the distribution of the coefficients of the impact of the Must Query PDMPs on labor force participation rates across different levels of minimum wage bindingness for various periods relative to treatment.

% CHECKED
In the upper right panel, we see the immediate impact of the policy.
For counties where the minimum wage is less binding, the effect is positive (1.37 p.p.) and significant (0.513, 2.22) at conventional levels.
The effects become smaller and more uncertain as the minimum wage becomes more binding.
In counties where the minimum wage has a higher impact on labor demand, the point estimate becomes negative (-0.693) but not significant (-2.54, 1.16).

% CHECKED
The effect appears to be persistent in the short run (upper right and lower right panels) but fades away after two years (lower right)\footnote{The confidence interval for the most binding counties is not displayed in this last panel for visualization purposes. The values are (-3.10, 1.90)}.
Figure 3 shows the time averages of the coefficients for two years after the law is passed.
The results remain positive and significant for the counties where the minimum wage is less binding (1.06 p.p., (0.584, 1.55)), while they are not significant for those where the minimum wage was more binding.

% CHECKED
The results seem to be driven primarily by the impact of the policy on prescription rates.
Although the overlap between the passing of the policies (Table 1) and the data for prescription rates (2006-2014) is limited, we can still obtain some estimates of the impact of Must Query PDMPs on prescription rates per capita.
Figure 4 runs the model (10) with prescription rates per capita as the dependent variable, showing the two months after the treatment is passed.
The policy is effective in decreasing the amount of prescriptions, with the biggest drops occurring where the minimum wage is the least and the most binding (the effects in these areas are not statistically different).
Although the target of these policies was not to decrease prescriptions \textit{per se}, but to tackle misuse cases, the results suggest that the policies are effective in reducing the overall amount of opioids prescribed.

% CHECKED
Moreover, this seems to be the main channel through which the policies affect labor market outcomes.
Figure 8 in the appendix shows that the relation between labor market outcomes and minimum wage bindingness is not significant when controlling for prescription rates.
Figures 9 and 10 in the appendix show the point estimates of the binscatter for different Kaitz indices for six months and two years after the policy is passed.
The results do not change much across different measures, especially the closer in time the treatment was implemented.

@@

% CHECKED
The analysis is repeated using the unemployment rate as the dependent variable.
The results are displayed in the same manner in Figure 5.
The distribution of the coefficients reveals that almost all the positive effects found in the labor force participation rates translated into an increase in the unemployment rate.
For counties where the minimum wage is less binding, the effect is positive (1.02 p.p.) and significant (1.46, 0.575) at conventional levels. 
The point estimate is only 0.35 p.p. lower than the one found for the labor force participation rate.
Although a positive effect is also found for counties where the minimum wage is more binding (0.640 p.p.), the effect is not significant (-0.0252, 1.31).

% CHECKED
These results shouldn't be surprising, at least in the short run: if people with severe opioid addictions start looking for a job after the implementation of a PDMP, they might not be able to find one immediately even in counties where the minimum wage is less binding.
Strikingly though, the impact on unemployment persists after two years.
The lower right panel shows that, even after the increase in labor force participation rates fades away, the increase in unemployment rates remains high (2.77 p.p.) and significant (1.87, 3.68) where the minimum wage is less binding.
Whether the causes of this persistence are related to opioid consumption (through, say, relapsing) or to characteristics of the workers or the labor market (e.g., if these workers face more frictional unemployment) remains an open question.

% CHECKED
The time average of the coefficients, displayed in Figure 6, confirm our previous result.
The effect of the policy on unemployment rates is positive and significant for counties where the minimum wage is less binding (1.25 p.p., (0.97, 1.52)), while it is not significant for those where the minimum wage is more binding (0.025 p.p., (-0.11, 0.16)).
Again, the results seem to be driven by the impact of the policy on prescription rates, especially in the medium run: Figure 11 in the appendix shows that the impact stops being significantly different from zero after one year once we control for prescriptions per capita.
Figures 12 and 13 in the appendix show the binscatter for different Kaitz indices for six months and two years after the policy is passed.
The results do not change substantially across different measures.

@@

@@

% CHECKED
Finally, and in accordance with the model outlined above, I present evidence that the rollout of Must Query PDMPs has had a positive impact on the consumption of illegal opioids in counties where the minimum wage was more binding.
For this purpose, I apply the same estimation strategy as before, but use the number of overdose deaths per capita for different opioids as the dependent variable. 
The data is at the state-year level, and the number of total poisoning deaths is not used as a control variable.
Figure 7 displays the scatter plot of the time average of the coefficients on the Kaitz-0.10 index for the states, for the two years after the policy is passed.
The opioids included are heroin, methadone, other opioids, and other synthetic narcotics, following the ICD-10 classification.

% CHECKED
The coefficients are regressed on the Kaitz index and a constant. 
Tables 2-5 show the values of the regression coefficients for each of the different opioids.
Although the evidence presented here is admittedly weaker than in previous sections due to data constraints, it is interesting that the only marginally significant relation is found for heroin.
This suggests that the positive relation between black market switching and low labor demand is present, but we cannot assert this relationship with the same degree of certainty.

% CHECKED
It seems unlikely that part of the effect is absorbed by the consumption of other opioids.
Methadone is often used as a treatment for opioid addiction and can be legally used under a doctor's supervision.
Other opioids include drugs such as morphine and codeine, but also semi-synthetic opioid analgesics (e.g., oxycodone, hydrocodone, hydromorphone, and oxymorphone), some of which are targeted by the PDMPs.
Finally, although the category of other synthetic narcotics includes fentanyl and its analogues, its use was not as widespread during the period considered as it is today.

@@

% CHECKED
This paper contributes to the understanding of how labor market regulations, particularly minimum wage rates, shape the relationship between opioid misuse and labor market outcomes.
By leveraging the staggered rollout of Must Query Prescription Drug Monitoring Programs (PDMPs) across states, this paper provides evidence of how labor market opportunities influence the incentives for opioid users to quit or switch to illegal alternatives.

% CHECKED
The findings reveal that labor force participation and unemployment rates increased as a result of prescription policies only in areas where the minimum wage was not binding.
In contrast, a surge in heroin consumption occurred where the minimum wage was binding.
These results underscore the importance of considering labor market regulations when evaluating the economic impacts of the opioid crisis, as they can generate heterogeneities often overlooked in the literature.
Hence, this study adds to the growing body of research that examines the economic effects of the opioid epidemic, emphasizing the need for a comprehensive approach that considers the incentives faced by addicted individuals in different labor market contexts.

% CHECKED
Future research should address some limitations of this study.
First, more detailed data on illegal opioid consumption and effective minimum wages (which can vary in metropolitan areas) could provide more precise estimates of the effects of labor market regulations on the opioid crisis's impact on labor outcomes.
More importantly, the mechanisms through which these regulations affect the incentives of addicted individuals should be further explored.
The relevance of minimum wage rates hinges on their impact on labor demand, so further exploration of this mechanism, including the impact of these policies on vacancies, would be valuable.
Finally, a more dynamic model of workers' incentives might account for the short-run effects on labor rates participation rates and the persistent increase in unemployment rates observed.

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_lab_for_rate.png}
    \caption{Results for (10) using labor force participation rates as the dependent variable. The bins are computed optimally following Cattaneo et al. (2024). A confidence interval is displayed around the point estimates, at 95\% level.}
    \label{fig:lab_for_rate}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_lab_for_rate_ta.png}
    \caption{Time average of the coefficients of Figure 2, for two years after the policy is passed. The bins are computed optimally following Cattaneo et al. (2024). A confidence interval is displayed around the point estimates, at 95\% level.}
    \label{fig:lab_for_rate_ta}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_prescriptions.png}
    \caption{Distribution of the unit-level effects using prescriptions per capita as the dependent variable. The bins are computed optimally following Cattaneo et al. (2024). A confidence interval is displayed around the point estimates, at 95\% level.}
    \label{fig:prescriptions}
\end{figure}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=0.8\textwidth]{pmq10_unemp_rate.png}
    \caption{Results for (10) using unemployment rates as the dependent variable. The bins are computed optimally following Cattaneo et al. (2024). A confidence interval is displayed around the point estimates, at 95\% level.}
    \label{fig:unemp_rate}
\end{figure}
